Estimating the Initial Susceptible Fraction in Dengue Dynamics

### 2014-09-22 draft_lm1.tex
updated the introduction

\documentclass[10pt]{article}
% amsmath package, useful for mathematical formulas
\usepackage{amsmath}
% amssymb package, useful for mathematical symbols
\usepackage{amssymb}
% cite package, to clean up citations in the main text. Do not remove.
\usepackage{cite}
\usepackage{hyperref}
% line numbers
\usepackage{lineno}
% ligatures disabled
\usepackage{microtype}
\DisableLigatures[f]{encoding = *, family = * }
% rotating package for sideways tables
%\usepackage{rotating}
% If you wish to include algorithms, please use one of the packages below. Also, please see the algorithm section of our LaTeX guidelines (http://www.plosone.org/static/latexGuidelines) for important information about required formatting.
%\usepackage{algorithmic}
%\usepackage{algorithmicx}
% Use doublespacing - comment out for single spacing
%\usepackage{setspace} 
%\doublespacing
% Text layout
\topmargin 0.0cm
\oddsidemargin 0.5cm
\evensidemargin 0.5cm
\textwidth 16cm 
\textheight 21cm
% Bold the 'Figure #' in the caption and separate it with a period
% Captions will be left justified
\usepackage[labelfont=bf,labelsep=period,justification=raggedright]{caption}
% Use the PLoS provided BiBTeX style
\bibliographystyle{plos2009}
% Remove brackets from numbering in List of References
\makeatletter
\renewcommand{\@biblabel}[1]{\quad#1.}
\makeatother

@@

% Title must be 150 characters or less
\begin{flushleft}
{\Large
\textbf{Estimating the Initial Susceptible Fraction in Dengue Dynamics}
}
% Insert Author names, affiliations and corresponding author email.
\\
Flavio Code\c{c}o Coelho$^{1\ast}$, 
Luiz Max de Carvalho$^{2}$, 
\\
\bf{1} Escola de Matem\'atica Aplicada , Funda\c{c}\~ao Get\'ulio Varga (FGV), Rio de Janeiro -- RJ, Brazil.
\\
\bf{2} Programa de Computa\c{c}\~ao Cient\'ifica (PROCC), Funda\c{c}\~ao Oswaldo Cruz, Rio de Janeiro -- RJ, Brazil.
\\
$\ast$ E-mail: fccoelho@fgv.br
\end{flushleft}

@@

\section*{Introduction}
% You may title this section "Methods" or "Models". 
% "Models" is not a valid title for PLoS ONE authors. However, PLoS ONE
% authors may use "Analysis" 
\section*{Methods}
\subsection*{The model}
\subsection*{Parameter estimation using Bayesian melding}
% Results and Discussion can be combined.
\section*{Results}
% We only support three levels of headings, please do not create a heading level below \subsubsection.
%\subsection*{Subsection 1}
\section*{Discussion}
% Do NOT remove this, even if you are not including acknowledgments.
\section*{Acknowledgments}
LMC is grateful to Dr. Leonardo Bastos for useful discussions on the estimation of $R(t)$.
\section*{References}
\bibliography{lm1}
\section*{Figure Legends}
% This section is for figure legends only, do not include
% graphics in your manuscript file.
%
%\begin{figure}
%\caption{
%{\bf Bold the first sentence.}  Rest of figure caption.  
%}
%\label{Figure_label}
%\end{figure}

>>>>

\documentclass[10pt]{article}
% amsmath package, useful for mathematical formulas
\usepackage{amsmath}
% amssymb package, useful for mathematical symbols
\usepackage{amssymb}
% cite package, to clean up citations in the main text. Do not remove.
\usepackage{cite}
\usepackage{hyperref}
% line numbers
\usepackage{lineno}
% ligatures disabled
\usepackage{microtype}
\usepackage{todonotes}
\DisableLigatures[f]{encoding = *, family = * }
% rotating package for sideways tables
%\usepackage{rotating}
% If you wish to include algorithms, please use one of the packages below. Also, please see the algorithm section of our LaTeX guidelines (http://www.plosone.org/static/latexGuidelines) for important information about required formatting.
%\usepackage{algorithmic}
%\usepackage{algorithmicx}
% Use doublespacing - comment out for single spacing
%\usepackage{setspace} 
%\doublespacing
% Text layout
\topmargin 0.0cm
\oddsidemargin 0.5cm
\evensidemargin 0.5cm
\textwidth 16cm 
\textheight 21cm
% Bold the 'Figure #' in the caption and separate it with a period
% Captions will be left justified
\usepackage[labelfont=bf,labelsep=period,justification=raggedright]{caption}
% Use the PLoS provided BiBTeX style
\bibliographystyle{plos2009}
% Remove brackets from numbering in List of References
\makeatletter
\renewcommand{\@biblabel}[1]{\quad#1.}
\makeatother

@@

% Title must be 150 characters or less
\begin{flushleft}
{\Large
\textbf{Estimating the Initial Susceptible Fraction in Dengue Dynamics}
}
% Insert Author names, affiliations and corresponding author email.
\\
Flavio Code\c{c}o Coelho$^{1\ast}$, 
Luiz Max de Carvalho$^{2}$, 
\\
\bf{1} Escola de Matem\'atica Aplicada , Funda\c{c}\~ao Getulio Vargas (FGV), 
Rio de Janeiro -- RJ, Brazil.
\\
\bf{2} Programa de Computa\c{c}\~ao Cient\'ifica (PROCC), Funda\c{c}\~ao Oswaldo Cruz, Rio de Janeiro -- RJ, Brazil.
\\
$\ast$ E-mail: fccoelho@fgv.br
\end{flushleft}

@@

\section*{Introduction}
When modeling infecctious disease spread using dynamic models, the initial 
conditions of the system, although key to match the simulated dynamics to 
observed data, are frequently overlooked. If the model is about a disease 
invading a new population one can simply assume the vast majority of the 
population to be susceptible. However for a disease which is well in its way to 
endemicity but not yet stable, it can be very hard to determine the 
immunological structure of the population without resorting to expensive large 
scale prevalence tests.

In the particular case of Dengue fever in Brasil, with four circulating 
serotypes, it can be very challenging to parameterize a simulation starting at 
any time other than when the date the disease was reintroduced in the country 
in the late 1980s. For SIR-like models, the relevant initial values are the 
number of infectious and susceptible  individuals. The number of infectious 
individuals are less challenging to determine, because the get very small 
during the winters and can be derived from incidence data  with is monitored 
continuously. The real challenge is to determine the number of susceptibles to 
a particular strain right at the beginning of the dengue season.

Multiple factors contribute to the dificulty in estimating the number of 
susceptibles in the population, first and foremost is the fact that dengue 
confers permanente immunity within serotypes but with only short-lived 
cross-immunity. Secondly, there are a larger number of asymptomatic cases in 
any epidemic, which nevertheless acquire immunity. Lastly Demography and 
migrations affect the number of susceptible in ways which are not easy to 
determine.

The initial number of susceptibles -- $S_0$ -- is a a very powerful parameter 
with the regards to epidemic risk. Its value affects directly the local $R_0$, 
thus being intrumental in predicting epidemics, but it also affects the speed 
and size of the epidemic.

In this paper, we estimate the initial number of susceptibles right before 
every epidemic in the last 18 years. For this we use a simplified model of 
dengue transmission in which the vector is left out. We also model the mix of 
circulating serotypes without discriminating between them. We then fit the 
model output to observed data to estimate the $S_0$ before every 
recorded epidemic in the Brazilian city of Rio de Janeiro. Rio de Janeiro has 
bee through many Dengue epidemics of various serotypes and sizes.

\section*{Methods}
\subsection*{The data}
The Data used to fit the model consists of weekly incidence from 1996 to 2014 
in the city of Rio de Janeiro, Brazil. From this data, we calculated series 
effective reproduction number, $R_t$, time series using the method described by 
Nishiura et al.\cite{nishiura}.\todo[inline]{Add here more information about 
the calculation of Rt such as generation time used, moving windows, etc}
\subsection*{The model}
\subsection*{Parameter estimation using Bayesian melding}
% Results and Discussion can be combined.
\section*{Results}
% We only support three levels of headings, please do not create a heading level below \subsubsection.
%\subsection*{Subsection 1}
\section*{Discussion}
% Do NOT remove this, even if you are not including acknowledgments.
\section*{Acknowledgments}
LMC is grateful to Dr. Leonardo Bastos for useful discussions on the estimation of $R(t)$.

\bibliography{lm1}
\section*{Figure Legends}
% This section is for figure legends only, do not include
% graphics in your manuscript file.
%
%\begin{figure}
%\caption{
%{\bf Bold the first sentence.}  Rest of figure caption.  
%}
%\label{Figure_label}
%\end{figure}
